---
title: "Beyond pairwise correlation: capturing nonlinear and higher-order dependence with distance statistics"
authors:
  - name: Benjamin Avanzi
    affiliation: Centre for Actuarial Studies, Department of Economics, University of Melbourne, Australia
  - name: Guillaume Boglioni Beaulieu
    affiliation: School of Risk and Actuarial Studies, University of New South Wales, Sydney, NSW 2052, Australia
  - name: Pierre Lafaye de Micheaux
    affiliation: School of Mathematics and Statistics, University of New South Wales, Sydney, NSW 2052, Australia
  - name: Ho Ming Lee
    affiliation: 
      - Centre for Actuarial Studies, Department of Economics, University of Melbourne, Australia
      - Faculty of Economics and Business, KU Leuven, Belgium
    corresponding: true
  - name: Bernard Wong
    affiliation: School of Risk and Actuarial Studies, University of New South Wales, Sydney, NSW 2052, Australia
  - name: Rui Zhou
    affiliation: Centre for Actuarial Studies, Department of Economics, University of Melbourne, Australia
bibliography: references.bib
format: 
  html:
    code-fold: false
execute:
  echo: true
engine: knitr
---


```{r, echo = FALSE}
# Loading required libraries
library(ggplot2)
library(patchwork)
library(dplyr)
library(HellCor)
library(TOSTER)
library(dcov)
```

```{r helper_functions}
get_pmat_corr <- function(dat, method = c("pearson", "spearman", "kendall"),
                          R = 9999) {
  # This function is used to get the p-values for testing correlation significance
  # Input: a dataframe with all numerical columns
  # output: p-value matrix of the choice of correlation
  method <- match.arg(method)
  
  dat <- as.data.frame(dat)
  pmat <- matrix(NA_real_, ncol(dat), ncol(dat))
  colnames(pmat) <- rownames(pmat) <- colnames(dat)
  
  for (i in 1:(ncol(dat) - 1)) {
    for (j in (i + 1):ncol(dat)) {
      x <- dat[[i]]
      y <- dat[[j]]
      
      keep <- complete.cases(x, y)
      x <- x[keep]
      y <- y[keep]
      
      pval <- if (method == "pearson") {
        TOSTER::boot_cor_test(x, y, method = "pearson", null = 0, alternative = "two.sided", R = R)$p.value
      } else {
        cor.test(x, y, method = method)$p.value
      }
      
      pmat[i, j] <- pval
      pmat[j, i] <- pval
    }
  }
  
  diag(pmat) <- NA
  pmat
}
```

# Introduction
This is the code for the paper ``Beyond pairwise correlation: capturing nonlinear and higher-order dependence with distance statistics" by Benjamin Avanzi, Guillaume Boglioni Beaulieu, Pierre Lafaye de Micheaux, Ho Ming Lee, Bernard Wong, and Rui Zhou.

## Motivating examples
This section provide the codes for generating the figures and statistics in the introduction of the paper, as well as the data generation process for simulated data.

### 2020 birth and death rates
Our first motivating example involves the birth and death rate data from different countries in the first trimester of 2020 [@CIA_WorldFactbook_worlddemographics]. This example is to illustrate that correlations are inherently for linear and monotonic relationships and hence a 0 correlation does not mean independence, and a small value does not necessarily means the actual dependence relationship is weak.

The birth and death rates are measured as the numbers of births and deaths per 1000 individuals, respectively, across 229 countries. The data can be loaded through the `HellCor` package in `R`.
```{r, birth_vs_death_rate_plot}
# Load birth vs death rate data
data("worlddemographics")

# Transform the data into ranks
worlddemographics_rank <- worlddemographics |>
  transform(
    FDeath = rank(Death.Rate.Pop) / (nrow(worlddemographics) + 1),
    FBirth = rank(Birth.Rate.Pop) / (nrow(worlddemographics) + 1)
  )

# Plotting both the raw data (p1) and the ranks (p2) out
p1 <- ggplot(worlddemographics, aes(x = Death.Rate.Pop, y = Birth.Rate.Pop)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Death rate per 1000 individuals",
    y = "Birth rate per 1000 individuals"
  ) +
  theme_minimal()

p2 <- ggplot(worlddemographics_rank, aes(x = FDeath, y = FBirth)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Rank of death rate per 1000 individuals",
    y = "Rank of birth rate per 1000 individuals"
  ) +
  theme_minimal()

p1 + p2
```

The corresponding correlation coefficients, and the corresponding $p$-values with the null hypothesis 
$$
\begin{aligned}
H_0 &: \theta=0,\\
H_1 &: \theta\neq0,
\end{aligned}
$$
for $\theta\in\{r \text{ (Pearson)}, \rho \text{ (Spearman)}, \tau \text{ (Kendall)}\}$ are given by
```{r, warning = FALSE, birthvdeath_compute_corr}
# Getting raw data for birth and death
birth_raw <- worlddemographics$Birth.Rate.Pop
death_raw <- worlddemographics$Death.Rate.Pop

set.seed(123) # This is for bootstrap test for correlation
cor_matrix_birth_death <- rbind(
  correlation = c(
    cor.test(birth_raw, death_raw, method = "pearson")$estimate,
    cor.test(birth_raw, death_raw, method = "spearman")$estimate,
    cor.test(birth_raw, death_raw, method = "kendall")$estimate
  ),
  p_value = c(
    # p-val for Pearson's corr needs boostrap here
    # As the null of the t-test assumed bivariate normal,
    # which is not necessarily true in this case.
    TOSTER::boot_cor_test(birth_raw, death_raw)$p.value, 
    cor.test(birth_raw, death_raw, method = "spearman")$p.value,
    cor.test(birth_raw, death_raw, method = "kendall")$p.value
  )
)

colnames(cor_matrix_birth_death) <- c("Pearson", "Spearman", "Kendall")
cor_matrix_birth_death
```

### Multi-line example
The other example is used to illustrate that the correlation is inherently pairwise, and it cannot detect association when the underlying association exist when they considered jointly. For this example, there are two business lines where we have the motor line with the following components:
\begin{equation*}
      \mathbf{X}=(X_1,X_2,X_3)=\text{(vehicle repair cost, bodily injury liability, claims handling cost)},
\end{equation*}
and $Y$ denotes the medical claim cost recorded under another line of business. The data is simulated using the following code
```{r, multi_line_data_generation}
set.seed(123)
n <- 5000

U1 <- runif(n) # latent for vehicle repair
U2 <- runif(n) # latent for bodily injury liability
U3 <- runif(n) # latent for claims handling

# latent for medical claims - it depends on both 
# vehicle damage and bodily injury
C <- (U1 + U2) %% 1 

# Independently distributed in Gamma
X1 <- qgamma(U1, shape = 2.5, scale = 2000)
X2 <- qgamma(U2, shape = 3.0, scale = 3000)

# Gamma as well, but dependends on (X1, X2)
Y <- qgamma(C, shape = 2.2, scale = 2500)

# Here we have claims handling cost, and there are two types of claim:
# Severity driven: handling cost goes up with medical cost (C)
# Review driven: handling cost higher for smaller medical cost (1-C)
p <- 0.50
R <- rbinom(n, size = 1, prob = p)
W <- ifelse(R == 1, C, 1 - C)

X3 <- qgamma(W, shape = 2.0, scale = 1200)

multi_line_dat <- data.frame(X1,X2,X3,Y)
```
Below we apply some exploratory data analysis. First, we look at the distribution of each of the claim costs. As expected, all of the shape looks like a $\Gamma$-distribution:
```{r, multi_line_univariate_plots}
par(mfrow = c(2,3))

# Histogram for each of the claim costs
hist(multi_line_dat$X1, main = "X1: Vehicle damage", xlab = "Loss", col = "grey")
hist(multi_line_dat$X2, main = "X2: Bodily injury", xlab = "Loss", col = "grey")
hist(multi_line_dat$X3, main = "X3: Claims handling", xlab = "Loss", col = "grey")
hist(multi_line_dat$Y, main = "Y: Medical claim", xlab = "Loss", col = "grey")

# Boxplot for log claim components
boxplot(log(multi_line_dat[, c("X1","X2","X3","Y")]),
        main = "Log-scale boxplots of claim components")
```
Then, we apply bivariate analysis. First, we plot the scatter plots for each pair of the variables using their ranks. Here ranks are used to reveal the dependence structure between the pairs.
```{r, multi_line_cross_plot}
par(mfrow = c(2,2))
col_group <- ifelse(R == 1, "red", "blue")
plot(rank(multi_line_dat$X1), rank(multi_line_dat$Y), pch = 16, cex = 0.35,
     xlab = "Rank of X1", ylab = "Rank of Y",
     main = "Medical claim (Y) vs vehicle damage (X1)")

plot(rank(multi_line_dat$X2), rank(multi_line_dat$Y), pch = 16, cex = 0.35,
     xlab = "Rank of X2", ylab = "Rank of Y",
     main = "Medical claim (Y) vs bodily injury (X2)")

plot(rank(multi_line_dat$X2), rank(multi_line_dat$X3), pch = 16, cex = 0.35,
     xlab = "Rank of X2", ylab = "Rank of X3",
     main = "Bodily injury (X2) vs vehicle damage (X2)")

plot(rank(multi_line_dat$X3), rank(multi_line_dat$Y), pch = 16, cex = 0.35,
     col = col_group, xlab = "Rank of X1", ylab = "Rank of Y",
     main = "Medical claim (Y) vs claims handling (X3)")
```
The correlation and the corresponding $p$-value matrices are given by below:
```{r, multi_line_corr}
cor(multi_line_dat) # Pearson correlation
get_pmat_corr(multi_line_dat, method = "pearson")

cor(multi_line_dat, method = "spearman") # Spearman correlation
get_pmat_corr(multi_line_dat, method = "spearman")

cor(multi_line_dat, method = "kendall") # Kendall correlation
get_pmat_corr(multi_line_dat, method = "kendall")
```
Here, as we can see the pairs $(X_1,X_2)$, $(X_1,Y)$, and $(X_2,Y)$ does not show any obvious pattern, suggesting independence. However, when we look at the scatter plot between $(X_1,X_2)$ coloured by $Y$, we can see that the value of $Y$ forms a diganoal pattern, i.e. knowing the value of $(X_1,X_2)$ provide information of $Y$, suggesting that they are not mutually independent. This raise one of the pitfall that correlation, and in general, thinking in pairwise dependence, can miss strutures where the dependence is considered jointly.
```{r, multi_line_X1X2Y_scatter}
# build eCDF for the multi-line data
multi_line_ECDF <- multi_line_dat %>%
  mutate(
    FX1 = rank(X1) / n(),
    FX2 = rank(X2) / n(),
    FX3 = rank(X3) / n(),
    FY = rank(Y) / n()
  )

ggplot(multi_line_ECDF, aes(x = FX1, y = FX2, colour = FY)) + 
  geom_point(size = 1.2, alpha = 0.8) + 
  scale_colour_gradient2(
    low = "blue",
    mid = "grey90",
    high = "red",
    midpoint = 0.5,
    name = expression(hat(F)[Y](Y))
  ) +
  labs(
    x = expression(hat(F)[X[1]](X[1])),
    y = expression(hat(F)[X[2]](X[2]))
  ) + 
  theme_minimal()
```

### Failure of CLT
Another important implication of relying on pairwise dependence statistics is that, many of the actuarial modelling assumption relies on the assumption that the random variables are (mutually) independent and identically distributed. It is possible to construct a sequence of identical and pairwise independent random variables but not mutually independent. However, this cannot be detected through bivariate dependence statistics such as the correlation. And it is possible that the central limit theorem (CLT) fails for such a sequence of random variables.

As illustrated by @avanzi2021counterexample and @beaulieu2021counterexamples, pairwise independence alone does not guarantee asymptotic normality. They construct a sequence $X_1,\ldots,X_n$ of pairwise independent, but not mutually independent, standard normal random variables such that the limiting distribution of
  \begin{equation}
      \sqrt{n}\,\frac{1}{n}\sum_{i=1}^n X_i
  \end{equation}
  coincides with that of the random variable
  \begin{equation}\label{eq:limit_dist}
      S=\sqrt{1-r^2}\,Z+r\chi,
  \end{equation}
where $Z\sim N(0,1)$, $\chi$ is a standardised chi-squared random variable independent of $Z$, and $0<r^2<1$. The limiting distribution does not necessarily converge to the normal distribution, as the CLT suggests.

```{r, failure_CLT}
set.seed(123)

# standardised chi-squared distribution with df = 1
rchi_std <- function(n){
  (rchisq(n, df = 1)-1)/sqrt(2)
}

rS <- function(n, r){
  sqrt(1 - r^2) * rnorm(n) + r * rchi_std(n)
}

r_vals <- c(0.6, 0.8, 0.95)
n <- 300000
x_grid <- seq(-4, 8, length.out = 1000)

samples <- lapply(r_vals, function(r) rS(n, r))
densities <- lapply(samples, density,
                    from = min(x_grid), to = max(x_grid),
                    n = 1000)

ymax <- max(
  sapply(densities, function(d) max(d$y)),
  max(dnorm(x_grid))
)
ymax <- 1.08 * ymax

par(mfrow = c(1,2), mar = c(5,5,3,1))
# ----- left panel: PDF -----
plot(densities[[1]], type = "l", lwd = 2, col = 2,
     main = "PDF", xlab = "x", ylab = "",
     xlim = range(x_grid), ylim = c(0, ymax))

for (i in 2:length(densities)) {
  lines(densities[[i]], lwd = 2, col = i + 1, lty = i)
}
lines(x_grid, dnorm(x_grid), lwd = 2, col = 1)

legend("topright",
       legend = c(paste0("r = ", r_vals), "N(0,1)"),
       col = c(2, 3, 4, 1),
       lty = c(1, 2, 3, 1),
       lwd = 2,
       bty = "n")

# ----- right panel: CDF -----
plot(x_grid, ecdf(samples[[1]])(x_grid), type = "l", lwd = 2, col = 2,
     main = "CDF", xlab = "x", ylab = "",
     xlim = range(x_grid), ylim = c(0, 1))

for (i in 2:length(samples)) {
  lines(x_grid, ecdf(samples[[i]])(x_grid),
        lwd = 2, col = i + 1, lty = i)
}
lines(x_grid, pnorm(x_grid), lwd = 2, col = 1)

legend("bottomright",
       legend = c(paste0("r = ", r_vals), "N(0,1)"),
       col = c(2, 3, 4, 1),
       lty = c(1, 2, 3, 1),
       lwd = 2,
       bty = "n")
```

## Illustration with Hellinger correlation
## Illustration with distance covariance/correlation

## Time series

```{r}
library(quantmod)
library(tseries)
getSymbols("^GSPC", src = "yahoo", from = "1926-01-01")
price <- Cl(GSPC)
ret <- monthlyReturn(price, type = "log")

adf.test(ret)
```


```{r}
library(dCovTS)
library(vars)
library(urca)

data(MortTempPart)

x <- MortTempPart[, c("cmort", "tempr", "part")]
x <- as.matrix(x)

sel <- VARselect(x, lag.max = 10, type = "const")
sel$criteria
sel$selection

fit_var <- VAR(x, p = 2, type = "const")
summary(fit_var)

E <- residuals(fit_var)


```

```{r}
# install.packages("tseries")

cmort_ts <- ts(MortTempPart$cmort) 
tempr_ts <- ts(MortTempPart$tempr) 
part_ts <- ts(MortTempPart$part) 

adf.test(cmort_ts)
adf.test(tempr_ts)
adf.test(part_ts)
```